---
title: "Pseudo mixture analysis"
author: "Xueyi Dong"
date: "14/06/2021"
output: 
  html_document: 
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(Rsubread)
library(limma)
library(edgeR)
library(ggplot2)
library(DESeq2)
library(EBSeq)
library(NOISeq)
```

```{r loadFC, echo=FALSE}
fc <- readRDS("counts_ONT.RDS")
fc2 <- readRDS("counts_ONT_mix.RDS")
```

```{r}
# Set output folder
OUT="./mix"
```

## DE analysis using limma

```{r}
x <- DGEList(counts = cbind(fc$counts[,1:6], fc2$counts))
#HCC827 proportion
x$samples$group <- rep(c("000", "100", "075", "050", "025"), rep(3, 5))
x$samples$mixprop <- rep(c(0, 1, 0.75, 0.5, 0.25), rep(3, 5))
x$samples$pair <- rep(c(1, 2, 3), 5)
x$genes <- fc$annotation
```

```{r}
# geneid <- strsplit2(rownames(x$counts), "\\.")[,1]
# genes <- select(Homo.sapiens, keys=geneid, columns=c("SYMBOL", "TXCHROM"), 
#                 keytype="ENSEMBL")
# x$genes <- genes[match(geneid, genes$ENSEMBL),]
```

```{r}
filter <- filterByExpr(x)
x <- x[filter, ,keep.lib.sizes=FALSE]
dim(x)

library(RColorBrewer)
col <- brewer.pal(5, "Set1")

x <- calcNormFactors(x)
cpm <- cpm(x, log=TRUE)
plotMDS(cpm, labels = x$samples$group, 
        col = rep(col, rep(3, 5)))

x$samples

human.gene <- grep("^ENSG", rownames(x))
sequin.gene <- grep("^R", rownames(x))
cpm.human <- cpm(x[human.gene, ], log=TRUE)
cpm.sequin <- cpm(x[sequin.gene, ], log=TRUE)
par(mfrow=c(1,2))
plotMDS(cpm.human, labels = x$samples$group, 
        col = rep(col, rep(3, 5)), main="Human genes")
plotMDS(cpm.sequin, labels = x$samples$group, 
        col = rep(col, rep(3, 5)), main = "Sequin genes")
```

### Make design and contrast

```{r}
design <- model.matrix(~0+x$samples$group+x$samples$pair)
colnames(design) <- gsub("x$samples$group", "g", colnames(design), fixed=TRUE)
colnames(design) <- gsub("x$samples$pair", "pair", colnames(design), fixed=TRUE)
design
contr <- makeContrasts(
  c100vs0=g100-g000,
  c75vs25=g075-g025,
  c50vs25=g050-g025,
  c75vs50=g075-g050,
  levels = colnames(design)
)
contr

anno <- read.table("/wehisan/home/allstaff/d/dong.x/annotation/sequins/rnasequin_genes_2.4.tsv", header = TRUE, stringsAsFactors = FALSE)
anno$logFC <- log(anno$MIX_B / anno$MIX_A)
```

### fit a separate model for sequins genes

```{r}
x.sequin <- x[sequin.gene,]
x.sequin <- calcNormFactors(x.sequin)
v.sequin <- voom(x.sequin, design = design, plot=TRUE)
fit.sequin <- lmFit(v.sequin)
fit.sequin <- contrasts.fit(fit.sequin, contrasts=contr)
efit.sequin <- eBayes(fit.sequin)
dt.sequin <- decideTests(efit.sequin)
summary(dt.sequin)
# tt.sequin <- topTable(efit.sequin, number = Inf)
```
```{r}
par(mfrow=c(2,2))
for(i in 1:4){
  plotMD(efit.sequin, status = dt.sequin[,i], column = i, values = c(-1, 1), col = c("blue", "red"), main=colnames(contr)[i])
}

```


### Fit a separate model for human genes

```{r}
x.human <- x[human.gene,]
x.human <- calcNormFactors(x.human)
v.human <- voom(x.human, design = design, plot=TRUE)
fit.human <- lmFit(v.human)
fit.human <- contrasts.fit(fit.human, contrasts=contr)
efit.human <- eBayes(fit.human)
dt.human <- decideTests(efit.human)
summary(dt.human)
# tt.human <- topTable(efit.human, number = Inf)
# plotMD(efit.human, status = dt.human[,2], column = 2, values = c(-1, 1), col = c("blue", "red"), main="Human genes")
```

Get DE genes

```{r}
DE.human.limma <- lapply(1:4, function(x){
  rownames(dt.human)[dt.human[,x] != 0]
})
DE.sequin.limma <- lapply(1:4, function(x){
  rownames(dt.sequin)[dt.sequin[,x] != 0]
})
```


```{r}
# write.table(tt.human, file.path(OUT, "topTableHuman.tsv"), sep="\t")
# write.table(tt.sequin, file.path(OUT, "topTableSequin.tsv"), sep="\t")
# write.table(tt.human[,c("GeneID","logFC", "adj.P.Val")], file.path(OUT, "resLimmaHuman.tsv"), sep = "\t", row.names = TRUE)
# write.table(tt.sequin[,c("GeneID", "logFC", "adj.P.Val")], file.path(OUT, "resLimmaSequin.tsv"), sep = "\t", row.names = TRUE)
```

### Fit non-linear model for the mixture series

```{r}
mix.fit.human <- fitmixture(v.human$E, mixprop = x$samples$mixprop)
```



## DE analysis using edgeR

### Human genes

```{r}
x.human <- estimateDisp(x.human, design)
plotBCV(x.human)
qlfit.human <- glmQLFit(x.human, design)
plotQLDisp(qlfit.human)
res.human <- lapply(colnames(contr), function(x){ glmQLFTest(qlfit.human, contrast=contr[,x])})
ttag.human <- lapply(res.human, function(x){topTags(x, n=Inf)})
# write.table(ttag.human[,c("GeneID","logFC", "FDR")], file.path(OUT, "resEdgerHuman.tsv"), sep="\t", row.names = TRUE)

DE.human.edgeR <- lapply(ttag.human, function(x){
  x$table$GeneID[x$table$FDR < 0.05]
})
```

```{r}
# plotMD(res.human, status = decideTestsDGE(res.human), values = c(1, -1), 
       # col = c("red", "blue"), legend = "topright")
```

### Sequin genes

```{r}
x.sequin <- estimateDisp(x.sequin, design)
plotBCV(x.sequin)
qlfit.sequin <- glmQLFit(x.sequin, design)
plotQLDisp(qlfit.sequin)
res.sequin <- lapply(colnames(contr), function(x){ glmQLFTest(qlfit.sequin, contrast=contr[,x])})
ttag.sequin <- lapply(res.sequin, function(x){topTags(x, n=Inf)})

# plotMD(res.sequin, status = decideTestsDGE(res.sequin), values = c(1, -1), 
       # col = c("red", "blue"), legend = "topright")
# write.table(ttag.sequin[,c("GeneID","logFC", "FDR")], file.path(OUT, "resEdgerSequin.tsv"), sep="\t", row.names = TRUE)

DE.sequin.edgeR <- lapply(ttag.sequin, function(x){
  x$table$GeneID[x$table$FDR < 0.05]
})
```

## DE analysis using DESeq2

```{r}
dds <- DESeqDataSetFromMatrix(countData=  cbind(fc$counts[,1:6], fc2$counts), 
                              colData = data.frame(
                                group = rep(c("000", "100", "075", "050", "025"), rep(3, 5))
                              ),
                              design = design)
keep <- rowSums(counts(dds)) > 1 & rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep,]
nrow(dds)

human.gene <- grep("^ENSG", rownames(dds))
sequin.gene <- grep("^R", rownames(dds))
dds.human <- dds[human.gene, ]
dds.sequin <- dds[sequin.gene, ]

dds.human <- DESeq(dds.human)
dds.sequin <- DESeq(dds.sequin)

deres.human <- lapply(colnames(contr), function(x){results(dds.human, contrast = contr[,x])})
deres.human <- lapply(deres.human, function(x){
  x$genes <- rownames(x)
  x})
deres.sequin <- lapply(colnames(contr), function(x){results(dds.sequin, contrast = contr[,x])})
deres.sequin <- lapply(deres.sequin, function(x){
  x$genes <- rownames(x)
  x
})

# write.table(res.human[,c("gene", "log2FoldChange", "padj")], file.path(OUT, "resDeseq2Human.tsv"), sep="\t", row.names = TRUE)
# write.table(res.sequin[,c("gene", "log2FoldChange", "padj")], file.path(OUT, "resDeseq2Sequin.tsv"), sep="\t", row.names = TRUE)

DE.human.DESeq2 <- lapply(deres.human, function(x){
  x$genes[x$padj < 0.05]
})
DE.sequin.DESeq2 <- lapply(deres.sequin, function(x){
  x$genes[x$padj < 0.05]
})
```

## DE analysis using EBSeq

```{r}
human.gene <- grep("^ENSG", rownames(fc$counts))
sequin.gene <- grep("^R", rownames(fc$counts))
counts.human <- cbind(fc$counts[human.gene, 1:6], fc2$counts[human.gene,])
counts.sequin <- cbind(fc$counts[sequin.gene, 1:6], fc2$counts[sequin.gene,])
Sizes.human = MedianNorm(counts.human)
Sizes.sequin = MedianNorm(counts.sequin)
Conditions = as.factor(rep(c("000", "100", "075", "050", "025"), rep(3, 5)))
comparisons = list(c("000", "100"), 
                          c("025", "075"),
                          c("025", "050"),
                          c("050", "075"))
EBOut.human = lapply(comparisons,
                     function(x){
                       sel = which(as.character(Conditions) %in% x)
                       out = EBTest(Data=counts.human[,sel], 
                                    Conditions = factor(Conditions[sel]),
                                    sizeFactors = Sizes.human[sel],
                                    maxround = 5)
                       out})
EBDERes.human = lapply(EBOut.human, function(x){GetDEResults(x, FDR = 0.05)})
EBOut.sequin = lapply(comparisons,
                     function(x){
                       sel = which(as.character(Conditions) %in% x)
                       out = EBTest(Data=counts.sequin[,sel], 
                                    Conditions = factor(Conditions[sel]),
                                    sizeFactors = Sizes.sequin[sel],
                                    maxround = 5)
                       out})
EBDERes.sequin = lapply(EBOut.sequin, function(x){GetDEResults(x, FDR = 0.05)})

GeneFC.human = lapply(EBOut.human, PostFC)
GeneFC.sequin = lapply(EBOut.sequin, PostFC)
# write.table(data.frame(
#   gene = rownames(EBDERes.human$PPMat)[match(names(GeneFC.human$PostFC), rownames(EBDERes.human$PPMat))],
#   logFC=log2(GeneFC.human$PostFC),
#   PPEE=EBDERes.human$PPMat[match(names(GeneFC.human$PostFC), rownames(EBDERes.human$PPMat)),1],
#   row.names = names(GeneFC.human$PostFC)
# ), file.path(OUT, "resEbseqHuman.tsv"), sep="\t", row.names = TRUE)
# write.table(data.frame(
#   gene = rownames(EBDERes.sequin$PPMat)[match(names(GeneFC.sequin$PostFC), rownames(EBDERes.sequin$PPMat))],
#   logFC=log2(GeneFC.sequin$PostFC),
#   PPEE=EBDERes.sequin$PPMat[match(names(GeneFC.sequin$PostFC), rownames(EBDERes.sequin$PPMat)),1],
#   row.names = names(GeneFC.sequin$PostFC)
# ), file.path(OUT, "resEbseqSequin.tsv"), sep="\t", row.names = TRUE)

DE.human.EBSeq <- lapply(EBDERes.human, function(x){x$DEfound})
DE.sequin.EBSeq <- lapply(EBDERes.sequin, function(x){x$DEfound})
```


## DE analysis using NOISeq

```{r}
# mylength <- fc$annotation[,6, drop=FALSE]
# human.gene <- grep("^ENSG", rownames(fc$counts))
# sequin.gene <- grep("^R", rownames(fc$counts))
# rownames(mylength) <- fc$annotation$GeneID
myfilt.human = filtered.data(counts.human, factor = x$samples$group, norm=FALSE, method = 1)
myfilt.sequin = filtered.data(counts.sequin, factor = x$samples$group, norm=FALSE, method = 1)
mydata.human <- readData(
  data = myfilt.human,
  # length = mylength[human.gene,, drop=FALSE],
  factors = data.frame(
    group = rep(c("000", "100", "075", "050", "025"), rep(3, 5))
  )
)
mydata.sequin <- readData(
  data = myfilt.sequin,
  # length = mylength[sequin.gene,, drop=FALSE],
  factors = data.frame(
    group = rep(c("000", "100", "075", "050", "025"), rep(3, 5))
  )
)

mynoiseqbio.human = lapply(comparisons, function(x){
  noiseqbio(mydata.human, norm = "tmm", factor = "group", conditions = x)
})
mynoiseqbio.sequin = lapply(comparisons, function(x){
  noiseqbio(mydata.sequin, norm = "tmm", factor = "group", conditions = x)
})

noires.human <- lapply(mynoiseqbio.human, function(x){x@results[[1]]})
noires.sequin <- lapply(mynoiseqbio.sequin, function(x){x@results[[1]]})
noires.human <- lapply(noires.human, function(x){
  x$FDR = 1 - x$prob
  x$gene = rownames(x)
  x})
noires.sequin <- lapply(noires.sequin, function(x){
  x$FDR = 1 - x$prob
  x$gene <- rownames(x)
  x})

# # write.table(noires.human[,c("gene", "log2FC", "FDR")], file.path(OUT, "resNoiseqHuman.tsv"), sep="\t", row.names = TRUE)
# write.table(noires.sequin[,c("gene", "log2FC", "FDR")], file.path(OUT, "resNoiseqSequin.tsv"), sep="\t", row.names = TRUE)
DE.human.NOISeq <- lapply(noires.human, function(x){
  x$gene[x$FDR < 0.05]
})
DE.sequin.NOISeq <- lapply(noires.sequin, function(x){
  x$gene[x$FDR < 0.05]
})
```

## Results

### Calculate recovery rate

```{r}
calcRecovery <- function(DE){
  tmp <- sapply(2:length(DE), function(x){
    sum(DE[[1]] %in% DE[[x]])
  }, simplify=TRUE)
  c(length(DE[[1]]), tmp)/length(DE[[1]])
}

rec.human <- as.data.frame(sapply(list(DE.human.limma, DE.human.edgeR, DE.human.DESeq2, DE.human.EBSeq, DE.human.NOISeq),
                    calcRecovery,
                    simplify = TRUE))
rownames(rec.human) <- c("100vs000", "075vs025", "050vs025", "075vs050")
colnames(rec.human) <- c("limma", "edgeR", "DESeq2", "EBSeq", "NOISeq")
library(data.table)
rec.human <- melt(rec.human, variable.name = "Method", value.name = "Recovery rate")
rec.human$Comparison <- factor(rep(c("100vs000", "075vs025", "050vs025", "075vs050"), 5), levels = c("100vs000", "075vs025", "050vs025", "075vs050"))

pdf(file.path(OUT, "humanRecovery.pdf"), height = 4)
ggplot(rec.human, aes(x=Comparison, y=`Recovery rate`, colour=Method, group=Method)) +
  geom_line(linetype=rep(c("solid", "longdash", "dashed", "solid", "solid"), rep(4, 5))) +
  geom_point(size=3, shape=rep(c(1, 2, 3, 4, 5), rep(4, 5))) +
  theme_bw() +
  scale_color_brewer(palette="Dark2")
dev.off()

rec.sequin <- as.data.frame(sapply(list(DE.sequin.limma, DE.sequin.edgeR, DE.sequin.DESeq2, DE.sequin.EBSeq, DE.sequin.NOISeq),
                    calcRecovery,
                    simplify = TRUE))
rownames(rec.sequin) <- c("100vs000", "075vs025", "050vs025", "075vs050")
colnames(rec.sequin) <- c("limma", "edgeR", "DESeq2", "EBSeq", "NOISeq")
library(data.table)
rec.sequin <- melt(rec.sequin, variable.name = "Method", value.name = "Recovery rate")
rec.sequin$Comparison <- factor(rep(c("100vs000", "075vs025", "050vs025", "075vs050"), 5), levels = c("100vs000", "075vs025", "050vs025", "075vs050"))

pdf(file.path(OUT, "sequinRecovery.pdf"), height = 4)
ggplot(rec.sequin, aes(x=Comparison, y=`Recovery rate`, colour=Method, group=Method)) +
  geom_line(linetype=rep(c("solid", "solid", "dashed", "solid", "solid"), rep(4, 5))) +
  geom_point(size=3, shape=rep(c(1, 2, 3, 4, 5), rep(4, 5))) +
  theme_bw() +
  scale_color_brewer(palette="Dark2")
dev.off()
```

### Power (number of DE genes)

```{r}
power.human <- as.data.frame(sapply(list(DE.human.limma, DE.human.edgeR, DE.human.DESeq2, DE.human.EBSeq, DE.human.NOISeq),
                    function(x){sapply(x, length, simplify = TRUE)},
                    simplify = TRUE))
rownames(power.human) <- c("100vs000", "075vs025", "050vs025", "075vs050")
colnames(power.human) <- c("limma", "edgeR", "DESeq2", "EBSeq", "NOISeq")
library(data.table)
power.human <- melt(power.human, variable.name = "Method", value.name = "Number of DE genes")
power.human$Comparison <- factor(rep(c("100vs000", "075vs025", "050vs025", "075vs050"), 5), levels = c("100vs000", "075vs025", "050vs025", "075vs050"))

pdf(file.path(OUT, "humanPower.pdf"), height = 4)
ggplot(power.human, aes(x=Comparison, y=`Number of DE genes`, colour=Method, group=Method)) +
  geom_line(linetype=rep(c("solid", "longdash", "dashed", "solid", "solid"), rep(4, 5))) +
  geom_point(size=3, shape=rep(c(1, 2, 3, 4, 5), rep(4, 5))) +
  theme_bw() +
  scale_color_brewer(palette="Dark2")
dev.off()
```

```{r}
power.sequin <- as.data.frame(sapply(list(DE.sequin.limma, DE.sequin.edgeR, DE.sequin.DESeq2, DE.sequin.EBSeq, DE.sequin.NOISeq),
                    function(x){sapply(x, length, simplify = TRUE)},
                    simplify = TRUE))
rownames(power.sequin) <- c("100vs000", "075vs025", "050vs025", "075vs050")
colnames(power.sequin) <- c("limma", "edgeR", "DESeq2", "EBSeq", "NOISeq")
library(data.table)
power.sequin <- melt(power.sequin, variable.name = "Method", value.name = "Number of DE genes")
power.sequin$Comparison <- factor(rep(c("100vs000", "075vs025", "050vs025", "075vs050"), 5), levels = c("100vs000", "075vs025", "050vs025", "075vs050"))

pdf(file.path(OUT, "sequinPower.pdf"), height = 4)
ggplot(power.sequin, aes(x=Comparison, y=`Number of DE genes`, colour=Method, group=Method)) +
  geom_line(linetype=rep(c("solid", "longdash", "dashed", "solid", "solid"), rep(4, 5))) +
  geom_point(size=3, shape=rep(c(1, 2, 3, 4, 5), rep(4, 5))) +
  theme_bw() +
  scale_color_brewer(palette="Dark2")
dev.off()
```

### Sequin FDR

```{r}
trueDE <- anno$NAME[anno$logFC != 0]
calcFDR <- function(DE){
  sapply(DE, function(x){
    FD = sum(!(x %in% trueDE))
    FD / length(x)
  }, simplify=TRUE)
}
                    
FDR.sequin <- as.data.frame(sapply(list(DE.sequin.limma, DE.sequin.edgeR, DE.sequin.DESeq2, DE.sequin.EBSeq, DE.sequin.NOISeq),
                     calcFDR, 
                     simplify = TRUE))

rownames(FDR.sequin) <- c("100vs000", "075vs025", "050vs025", "075vs050")
colnames(FDR.sequin) <- c("limma", "edgeR", "DESeq2", "EBSeq", "NOISeq")
FDR.sequin <- melt(FDR.sequin, variable.name = "Method", value.name = "False discovery rate")
FDR.sequin$Comparison <- factor(rep(c("100vs000", "075vs025", "050vs025", "075vs050"), 5), levels = c("100vs000", "075vs025", "050vs025", "075vs050"))

pdf(file.path(OUT, "sequinFDR.pdf"), height = 4)
ggplot(FDR.sequin, aes(x=Comparison, y=`False discovery rate`, colour=Method, group=Method)) +
  geom_line(linetype=rep(c("solid", "dashed", "solid", "longdash", "solid"), rep(4, 5))) +
  geom_point(size=3, shape=rep(c(1, 2, 3, 4, 5), rep(4, 5))) +
  theme_bw() +
  scale_color_brewer(palette="Dark2")
dev.off()
```

### Sequin TPR

```{r}
calcTPR <- function(DE){
  sapply(DE, function(x){
    TP = sum(x %in% trueDE)
    TP / length(trueDE)
  }, simplify=TRUE)
}
                    
TPR.sequin <- as.data.frame(sapply(list(DE.sequin.limma, DE.sequin.edgeR, DE.sequin.DESeq2, DE.sequin.EBSeq, DE.sequin.NOISeq),
                     calcTPR, 
                     simplify = TRUE))

rownames(TPR.sequin) <- c("100vs000", "075vs025", "050vs025", "075vs050")
colnames(TPR.sequin) <- c("limma", "edgeR", "DESeq2", "EBSeq", "NOISeq")
TPR.sequin <- melt(TPR.sequin, variable.name = "Method", value.name = "True positive rate")
TPR.sequin$Comparison <- factor(rep(c("100vs000", "075vs025", "050vs025", "075vs050"), 5), levels = c("100vs000", "075vs025", "050vs025", "075vs050"))

pdf(file.path(OUT, "sequinTPR.pdf"), height = 4)
ggplot(TPR.sequin, aes(x=Comparison, y=`True positive rate`, colour=Method, group=Method)) +
  geom_line(linetype=rep(c("solid", "dashed", "solid", "longdash", "solid"), rep(4, 5))) +
  geom_point(size=3, shape=rep(c(1, 2, 3, 4, 5), rep(4, 5))) +
  theme_bw() +
  scale_color_brewer(palette="Dark2")
dev.off()
```


## Session information

```{r}
sessionInfo()
```

